---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2026"
subtitle: "Assignment 2 - Due date 01/27/26"
author: "Abrao Soares Pereira"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp26.Rmd"). Then change "Student Name" on line 4 with your name.

Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai.


## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts=list(width.cutoff=80), tidy=TRUE)
```


```{r message= FALSE, warning=FALSE}
#Load required package
library(forecast)
library(ggplot2)
library(dplyr)
library(cowplot)
library(tseries)
library(readxl)
```

## Data set information

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2025 Monthly Energy Review. The spreadsheet is ready to be used. Refer to the file "M2_ImportingData_XLSX.Rmd" in our Lessons folder for instructions on how to read $.xlsx$ files.

```{r tidy=TRUE}
#Importing data set
energy_data <- read_excel(
  path = "~/Documents/ENVIRON 797 /Exercise_and_Assignments/Class_Exercise_and_Assignments/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
  skip = 12,
  sheet = "Monthly Data",
  col_names = FALSE
)

#Extract column names from row 11
read_col_names <- read_excel(
  path = "~/Documents/ENVIRON 797 /Exercise_and_Assignments/Class_Exercise_and_Assignments/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",
  skip = 10,
  n_max = 1,
  sheet = "Monthly Data",
  col_names = FALSE
)

#Assign column names
colnames(energy_data) <- as.character(unlist(read_col_names))

head(energy_data)

```

## Question 1

You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
#Select the columns for this analysis
energy_subset <- energy_data %>%
  select(`Total Biomass Energy Production`,
         `Total Renewable Energy Production`,
         `Hydroelectric Power Consumption`)

#Convert data to numeric 
energy_subset <- energy_subset %>%
  mutate(across(everything(), as.numeric))

head(energy_subset)
```

## Question 2

Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}

#The data starts on January 1973 
energy_ts <- ts(energy_subset, start = c(1973, 1), frequency = 12)
head(energy_ts)
```

## Question 3

Compute mean and standard deviation for these three series.

```{r}
#Compute mean and standard deviation for each of the three series
energy_mean <- apply(energy_ts, 2, mean, na.rm = TRUE)
energy_sd   <- apply(energy_ts, 2, sd,   na.rm = TRUE)
energy_stats <- data.frame(Mean = energy_mean, SD = energy_sd)
energy_stats

```

## Question 4

Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
#Total Biomass Energy Production Plot 
plot_biomass <- autoplot(energy_ts[,"Total Biomass Energy Production"]) +
  geom_hline(yintercept = energy_mean[1], color = "red", linetype = "dashed") +
  labs(title = "Total Biomass Energy Production (1973-2025)",
       x = "Year", y = "Trillion Btu") +
  theme_minimal()

#Total Renewable Energy Plot 
plot_renewable <- autoplot(energy_ts[,"Total Renewable Energy Production"]) +
  geom_hline(yintercept = energy_mean[2], color = "blue", linetype = "dashed") +
  labs(title = "Total Renewable Energy Production (1973-2025)",
       x = "Year", y = "Trillion Btu") +
  theme_minimal()

#Hydroelectric Power Consumption Plot
plot_hydro <- autoplot(energy_ts[,"Hydroelectric Power Consumption"]) +
  geom_hline(yintercept = energy_mean[3], color = "darkorange", linetype = "dashed") +
  labs(title = "Hydroelectric Power Consumption (1973-2025)",
       x = "Year", y = "Trillion Btu") +
  theme_minimal()

plot_biomass
plot_renewable
plot_hydro

```

> Answer: The biomass and renewable energy plots show a clear increasing trend, with most recent values sitting well above the mean. Hydroelectric power is different, it fluctuates around the mean with no clear long-term trend, but shows a repeating seasonal pattern. 

## Question 5

Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
#Pearson correlation among the three series
energy_corr <- cor(energy_subset, use = "complete.obs")
energy_corr

```

> Looking at the correlation matrix, biomass and renewables are strongly correlated, while hydro shows almost no relationship with either. The biomass-renewables correlation is statistically significant given the sample size. 


## Question 6

Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
#ACF up to lag 40 for the three series
par(mfrow = c(1,1))
Acf(energy_ts[,"Total Biomass Energy Production"],   lag.max = 40,
    main = "ACF - Total Biomass Energy Production")
Acf(energy_ts[,"Total Renewable Energy Production"], lag.max = 40,
    main = "ACF - Total Renewable Energy Production")
Acf(energy_ts[,"Hydroelectric Power Consumption"],   lag.max = 40,
    main = "ACF - Hydroelectric Power Consumption")


```

> The ACFs for Biomass and Renewables shows on the plot both slowly and stay positive on all the lags, which is a sign of a strong trend and non-stationarity. Hydro is different, its ACF oscillates with clear peaks at lags 12, 24, and 36, which reflects the yearly seasonality in hydropower.


## Question 7

Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
#PACF up to lag 40 for the three series
par(mfrow = c(1,1))
Pacf(energy_ts[,"Total Biomass Energy Production"],   lag.max = 40,
     main = "PACF - Total Biomass Energy Production")
Pacf(energy_ts[,"Total Renewable Energy Production"], lag.max = 40,
     main = "PACF - Total Renewable Energy Production")
Pacf(energy_ts[,"Hydroelectric Power Consumption"],   lag.max = 40,
     main = "PACF - Hydroelectric Power Consumption")
```

>  The PACFs show that most of the high correlations from Q6 collapse to just one or two significant lags. Lag 12 still stands out, especially for Hydro, which confirms the seasonal pattern. The main difference from Q6 is that the ACF reflects the total correlation including the indirect effect through earlier lags, while the PACF isolates the direct correlation at each lag.
